Fast Automatic Bayesian Cubature Using Lattice Sampling

### 2019-08-17 BayesianCubaturePaper2018-Rev1eRefs.tex
Updates from Overleaf

% This is a general template file for the LaTeX package SVJour3
% for Springer journals. Original by Springer Heidelberg, 2010/09/16
%
% Use it as the basis for your article. Delete % signs as needed.
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
\RequirePackage{amsmath}
\RequirePackage{fix-cm}
%
\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
%\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{graphicx}
\usepackage{amssymb}
%\usepackage{cite}
%
% insert here the call for the packages your document requires
%\usepackage{mathptmx}      % use Times fonts if available on your TeX system
%\usepackage{latexsym}
% etc.
%
% Jag's 
\usepackage{tabu}
\usepackage{cancel}
\usepackage{algorithm}
\usepackage{algorithmicx}
\usepackage{algpseudocode}
%\usepackage[caption=false]{subfig}
\usepackage{subcaption}
\usepackage{booktabs, mathtools}
\usepackage[dvipsnames]{xcolor}
\usepackage{url}
\usepackage[titletoc,title]{appendix}
%\usepackage{refcheck}  % points out unused labels, refs ...
\usepackage[htt]{hyphenat}

@@

% please place your own definitions here and don't use \def but
% \newcommand{}{}
\newtheorem{prop}{Proposition}
\providecommand{\HickernellFJ}{Hickernell}
\newcommand{\bm}[1]{\boldsymbol{#1}}
\newcommand{\mSigma}{\mathsf{\Sigma}}
\newcommand{\mB}{\mathsf{B}}
\newcommand{\smallocite}[1]{{\small\ocite{#1}}}
% \newcommand{\bm}[1]{\boldsymbol{#1}}
\newcommand{\dif}[1]{\text{d}{#1}}
\newcommand{\D}[1]{\text{d}{#1}}
\newcommand{\trace}[1]{\textup{trace}{#1}}

@@

The integral may often be expressed as
\begin{align}
\label{eqn:defn_mu}
\mu:= \Ex[f(\boldsymbol{X})] = \int_{[0,1]^d} f(\vx)\, \dif\vx, 
\end{align}
where $f:[0,1]^d \to \reals$ is the integrand, and $\boldsymbol{X} \sim \mathcal{U}[0,1]^d$.  The process of transforming the original integral into the form of \eqref{eqn:defn_mu} is not addressed here. See \cite[Section 2.11]{DicEtal14a} 
for a discussion of variable transformations. The cubature may be an affine function of integrand values:
\begin{align}
\label{eqn:defn_hmu}  % remove this
\hmu := w_0 + \sum_{i=1}^{n} f(\vx_i) w_i,
\end{align}
where the weights, $w_0$, and  $\vw = (w_i)_{i=1}^n \in \reals^n$, and the nodes, $\{\vx_i\}_{i=1}^n \subset [0,1]^d$, are chosen to make the error, $\abs{\mu - \hmu}$, small. The integration domain $[0,1]^d$ is convenient for the low discrepancy node sets \cite{DicEtal14a,SloJoe94} that we use.  The nodes are assumed to be deterministic.

Users of cubature algorithms typically want the error to be no greater than their specified error tolerance, denoted by $\varepsilon$.  That is, they want
\begin{align}
\label{eqn:err_crit} 
\abs{\mu - \hmu} \leq \errtol .
\end{align}
Some stopping criteria for choosing $n$ are heuristic.  Rigorous algorithms satisfying \eqref{eqn:err_crit}  typically require strong a priori assumptions about the integrand, such as an upper bound on its variance (for simple Monte Carlo) or total variation (for quasi-Monte Carlo).  We take a Bayesian approach by constructing a stopping criterion that is based on a credible interval.  We build upon the work of Briol et al.~\cite{BriEtal18a}, Diaconis~\cite{Dia88a}, O'Hagan~\cite{OHa91a}, Ritter~\cite{Rit00a}, Rasmussen and Ghahramani~\cite{RasGha03a}, and others.  Our algorithm is an example of \emph{probabilistic numerics}.

@@

Hickernell~\cite{Hic17a} compares different approaches to cubature error analysis depending on whether the rule is deterministic or random and whether the integrand is assumed to be deterministic or random.  Error analysis that assumes a deterministic integrand lying in a Banach space leads to an error bound that is typically impractical for deciding how large $n$ must be to satisfy \eqref{eqn:err_crit}.  The deterministic error bound includes a (semi-) norm of the integrand, often called the variation, which is often more complex to compute than the original integral.

Hickernell and Jim\'enez-Rugama~\cite{HicJim16a,JimHic16a} have developed stopping criteria for cubature rules based on low discrepancy nodes by tracking the decay of the discrete Fourier coefficients of the integrand.  The algorithm proposed here also relies on discrete Fourier coefficients, but in a different way.  Although we only explore automatic Bayesian cubature for absolute error tolerances, the recent work by Hickernell, Jim\'enez-Rugama, and Li~\cite{HicEtal17a} suggests how one might accommodate more general error criteria, such as relative error tolerances.

@@

We assume that the integrand, $f$, is an instance of a Gaussian stochastic process  i.e., $f \sim \mathcal{GP}(m,s^2 C_\vtheta)$  \cite{BriEtal18a,Dia88a,OHa91a,Rit00a,RasGha03a}.  Specifically, $f$ is a real-valued random function with constant mean $m$ and covariance kernel $s^2C_\vtheta$:
\begin{gather*}
        m = \Ex[f(\vx)] \qquad \forall \vx \in \reals^d, \\
        \Ex\{[f(\vt) - m][f(\vx) -m]\} = s^2 C_\vtheta(\vt,\vx) \qquad \forall \vt, \vx \in \reals^d.
\end{gather*}
Here $s$ is a positive scale factor, and $C_\vtheta: [0,1]^d \times [0,1]^d \to \mathbb{R} $ is a symmetric, positive-definite function and parameterized by the vector $\vtheta$:
\begin{multline} \label{FJH:eq:CondPosDef}
\mC_\vtheta^T = \mC_\vtheta,  \; \va^T \mC_\vtheta \va > 0,  \text{ where }  \mC_\vtheta = \left(  C_\vtheta(\vx_i,\vx_j)  \right)_{i,j=1}^n,\\
 \forall \va \ne 0, \;
 n\in \mathbb{N}, \; \text{ distinct }\vx_1, \ldots, \vx_n \in [0,1]^d.
\end{multline}
Procedures for estimating or integrating out the hyperparameters $m$, $s$, and $\vtheta$ are explained later in this section.

@@

\begin{lemma} \cite[(A.6), (A.11--13)]{RasWil06a} \label{thrm:condDist} If $\vY = (\vY_1, \vY_2)^T \sim \calN (\vm,\mSigma)$, where $\vY_1$ and $\vY_2$ are random vectors of arbitrary length, and 
	\begin{gather*}
	\vm = \begin{pmatrix} \vm_1 \\ \vm_2 \end{pmatrix} = \begin{pmatrix} \Ex(\vY_1) \\ \Ex(\vY_2) \end{pmatrix}, \\
	\mSigma = \begin{pmatrix}
	\mSigma_{11} & \mSigma_{21}^T \\ 	\mSigma_{21} & \mSigma_{22}
	\end{pmatrix} =
	\begin{pmatrix}
	\var(\vY_{1}) & \cov(\vY_{1}, \vY_2) \\ 	\cov(\vY_2,\vY_{1}) & \var(\vY_{2}),
	\end{pmatrix} 
	\end{gather*}
	then 
	\begin{align*}
	\vY_1 \vert \vY_2 \; \sim \; \calN \bigl(\vm_1 + \mSigma_{21}^T \mSigma_{22}^{-1}(\vY_2 - \vm_2),  \; \mSigma_{11} - \mSigma_{21}^T \mSigma_{22}^{-1} \mSigma_{21} \bigr).
	\end{align*}
Moreover, the inverse of the matrix $\mSigma$ may be partitioned as
\begin{gather*}
\mSigma^{-1} = \begin{pmatrix} \mA_{11} & \mA_{21}^T \\ \mA_{21} & \mA_{22} \end{pmatrix}, \\
\mA_{11} = (\mSigma_{11} - \mSigma_{12} \mSigma_{22}^{-1} \mSigma_{21})^{-1}, \qquad 
\mA_{21} = -  \mSigma_{22}^{-1} \mSigma_{21} \mA_{11}, \\ 
\mA_{22} = \mSigma_{22}^{-1} + \mSigma_{22}^{-1} \mSigma_{21} \mA_{11} \mSigma_{21}^T \mSigma_{22}^{-1}.
\end{gather*}

@@

Let $\mA = \mC^{-1}_\vtheta$, let $\vzeta = \mA (\vy - m \vone)$, and partition $\mC_\vtheta$, $\mA$, and $\vzeta$ as
\begin{equation*}
\mC_\vtheta = \begin{pmatrix} c_{ii}  & \vC_{-i,i}^T \\  \vC_{-i,i} & \mC_{-i,-i}\end{pmatrix}, \qquad
\mA = \begin{pmatrix} a_{ii}  & \vA_{-i,i}^T \\  \vA_{-i,i} & \mA_{-i,-i}\end{pmatrix}, \qquad \vzeta = \begin{pmatrix} \zeta_i   \\  \vzeta_{-i} \end{pmatrix},
\end{equation*}
where the subscript $i$ denotes the $i^{\text{th}}$ row or column, and the subscript $-i$ denotes all rows or columns except the $i^{\text{th}}$. Following this notation, Lemma \ref{thrm:condDist} implies that 
\begin{align*}
\mathring{y}_i & = m + \vC^T_{-i,i} \mC_{-i,-i}^{-1} (\vy_{-i} -m \vone)  \\
\zeta_i  & = a_{ii}(y_i - m) + \vA_{-i,i}^T(\vy_{-i} - m \vone) \\
& = a_{ii}[(y_i - m) - \vC^T_{-i,i} \mC_{-i,-i}^{-1} (\vy_{-i} -m \vone)] \\
& = a_{ii}(y_i - \mathring{y}_i).
\end{align*}
Thus, \eqref{FJH:eq:CVA} may be re-written as 
\begin{align*} %\label{FJH:eq:CVB}
\textup{CV} = \sum_{i=1}^n \left(\frac{\zeta_i}{a_{ii}} \right)^2, \qquad \vzeta = \mC^{-1}_\vtheta(\vy - m \vone).
\end{align*}
The \emph{generalized cross-validation} criterion (GCV) replaces the $i^{\text{th}}$ diagonal element of $\mA$ in the denominator by the average diagonal element of $\mA$ \cite{CraWah79a,GolHeaWah79a,Wah90}:
\begin{equation*} 
\textup{GCV}
= \frac{\sum_{i=1}^n\zeta_i^2}{\left(\frac 1n \sum_{i=1}^n a_{ii} \right)^2} 
= \frac{(\vy - m\vone)^T \mC^{-2}_\vtheta (\vy - m \vone)}{\left(\frac 1n \trace(\mC_\vtheta^{-1}) \right)^2}.
\end{equation*}

@@

Genz \cite{Gen93} introduced a variable transformation to transform \eqref{eqn:GaussDef} into an integral on the unit cube.  Not only does this variable transformation accommodate domains that are (semi-)infinite, it also tends to smooth out the integrand better, which expedites the cubature.  Let $\mSigma= \mL \mL^T$ be the Cholesky decomposition where $\mL = (l_{jk})_{j,k=1}^d$ is a lower triangular matrix.  Iteratively define
\allowdisplaybreaks
\begin{gather*}
\alpha_1 = \Phi(a_1), \qquad \beta_1 = \Phi(b_1),
\\
\alpha_j(x_1,\dots,x_{j-1}) = 
\Phi
\left(
\frac{1}{l_{jj}} 
\left(
a_j - \sum_{k=1}^{j-1} l_{jk} \Phi^{-1}(\alpha_k + x_k(\beta_k-\alpha_k))
\right)
\right), 
\\
\hspace{8cm} j=2,\dots,d,
\\
\beta_j(x_1,\dots,x_{j-1}) = 
\Phi
\left(
\frac{1}{l_{jj}} 
\left(
b_j - \sum_{k=1}^{j-1} l_{jk} \Phi^{-1}(\alpha_k + x_k(\beta_k-\alpha_k))
\right)
\right), 
\\
\hspace{8cm} j=2,\dots,d,
\end{gather*}
\begin{equation}
\label{fGenzdef}
f_{\text{Genz}}(\vx) = \prod_{j=1}^d [\beta_j(\vx) - \alpha_j(\vx)],
\end{equation}
where $\Phi$ is the univariate cumulative standard Gaussian distribution function.  Then, $\mu = \int_{[0,1]^{d'-1}} f_{\text{Genz}}(\vx) \, \dvx$. This approach transforms a $d'$ dimensional integral into a $d=d'-1$ dimensional integral.

@@

We use the following parameter values in the simulation: 
\begin{align*}
d' = 3, \quad \va = \begin{pmatrix}
-6 \\ -2 \\ -2
\end{pmatrix}, \quad 
\vb = \begin{pmatrix}
5 \\ 2 \\ 1
\end{pmatrix} , \quad 
\mL = \begin{pmatrix}
4 & 1 & 1 \\ 0 & 1 & 0.5 \\ 0 & 0 & 0.25
\end{pmatrix}.
\end{align*}
The node sets are randomly scrambled Sobol' points \cite{DicEtal14a,DicPil10a}.  The results are for 400 randomly chosen $\varepsilon$ in the interval $[10^{-5}, 10^{-2}]$ as shown in \figref{fig:MVN_Metern_d2b2}. In each run, the nodes are randomly scrambled.  The empirical Bayes credible intervals are used for stopping criteria.  We  observe that the algorithm meets the error criterion 95\% of the time even though we used 99\% credible intervals. One possible explanation is that the matrix inversions in the algorithm are ill-conditioned leading to numerical inaccuracies.  Another possible explanation is that this Mat\'ern covariance kernel is not a good match for the integrand.
%with Intel i7 3630QM and 16GB RAM memory
%On our test computer,

@@

An example of $64$ nodes is given in \figref{latticefig}.  The even coverage of the unit cube is ensured by a well chosen generating vector, $\vh$.  The choice of generating vector is typically done offline by computer search.  See \cite{DicEtal14a,HicNie03a} for more on extensible integration lattices.
\begin{figure}[htp]
	\centering
	\includegraphics[height=5cm]{ShiftedLatticePoints}
	\caption{Example of a shifted integration lattice node set  in $d=2$. 
	This figure can be reproduced using \code{PlotPoints.m} in GAIL} \label{latticefig}
\end{figure}

@@

A family of shift-invariant kernels is constructed via even degree Bernoulli polynomials:
\begin{multline}
\label{the_kernel_eqn_bernoulli}
K_\vtheta(\vx) =
\prod_{l=1}^d \biggl[
1 - (-1)^{r} \eta B_{2r}( x_l) \biggr], \\  
\forall \vt,\vx \in [0,1]^d, \  \vtheta = (r,\eta), \ r \in \naturals, \ \eta > 0.
\end{multline}
\end{subequations}
Symmetric, periodic, positive definite kernels of this form appear in  \cite{DicEtal14a,Hic96a}.  Bernoulli polynomials are described in \cite[Chapter 24]{OlvEtal10a}.

@@

Lattice cubature rules are known to have convergence rates that depend on the smoothness of the integrands, but that are rather independent of the choice of the integration lattice \cite{DicEtal14a}.  Thus, we expect integration lattice node sets to perform well regardless of the smoothness of the covariance kernel.  The bigger concern is whether the derivatives of the integrand are as smooth as the covariance kernel implies.  This topic is touched upon again in Section  \ref{period_var_tx}.

@@

We have implemented the fast adaptive Bayesian cubature algorithm in MATLAB as part of the Guaranteed Adaptive Integration Library (GAIL) \cite{ChoEtal17b} as \linebreak[4]  \code{cubBayesLattice\_g}. This algorithm uses the covariance kernel defined in \eqref{eqn:ourkernel} with  $r=1$ and $2$, and the periodizing variable transforms in  \secref{period_var_tx}.  The rank-1 lattice node generator is taken from \cite{Nuy17a}  (\code{exod2\_base2\_m20}).

@@

Three integrals were evaluated using the GAIL algorithm \code{cubBayesLattice\_g}:  a multivariate Gaussian  probability, Keister's example, and an option pricing example. 
These three integrands are defined below.
The sequences $\{\vx_i\}_{i=1}^\infty$ are the randomly shifted lattice node sequences supplied by GAIL. 
For each integral and each of our stopping criteria---empirical Bayes, full Bayes, and generalized cross-validation---our algorithm was run for $400$ different randomly chosen error tolerances. The error tolerances were chosen randomly in an interval depending on the difficulty of the problem. In each run, the nodes were also randomly shifted with $\mathcal{U}[0,1]$ shifts independent of each other and the error tolerances. The accuracy of the algorithm depends mildly on the shift; there is no universally optimal shift.
For each test, the execution times are plotted against $\abs{\mu - \hmu}/\varepsilon$.  We expect $\abs{\mu - \hmu}/\varepsilon$ to be no greater than one, but hopefully not too much smaller than one, which would indicate a stopping criterion that is too conservative.

@@

Figures \ref{fig:mvn-guaranteed-MLE} to \ref{fig:optprice-guaranteed-GCV} can be reproduced using the script \break \code{cubBayesLattice\_guaranteed\_plots.m} in GAIL.

@@

% BibTeX users please use one of
%\bibliographystyle{spbasic}      % basic style, author-year citations
%\bibliographystyle{spmpsci}      % mathematics and physical sciences
%\bibliographystyle{spphys}       % APS-like style for physics
%\bibliography{}   % name your BibTeX data base
\bibliography{FJHown23,FJH23}
\bibliographystyle{spmpsci}

>>>>

% This is a general template file for the LaTeX package SVJour3
% for Springer journals. Original by Springer Heidelberg, 2010/09/16
%
% Use it as the basis for your article. Delete % signs as needed.
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
\RequirePackage{amsmath}
\RequirePackage{fix-cm}
%
\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
%\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{xspace}
\usepackage{amssymb}
%\usepackage{cite}
%
% insert here the call for the packages your document requires
%\usepackage{mathptmx}      % use Times fonts if available on your TeX system
%\usepackage{latexsym}
% etc.
%
% Jag's 
\usepackage{tabu}
\usepackage{cancel}
\usepackage{algorithm}
\usepackage{algorithmicx}
\usepackage{algpseudocode}
%\usepackage[caption=false]{subfig}
\usepackage{subcaption}
\usepackage{booktabs, mathtools}
\usepackage[dvipsnames]{xcolor}
\usepackage{url}
\usepackage[titletoc,title]{appendix}
%\usepackage{refcheck}  % points out unused labels, refs ...

@@

% please place your own definitions here and don't use \def but
% \newcommand{}{}
\newtheorem{prop}{Proposition}
\providecommand{\HickernellFJ}{Hickernell\xspace}
\newcommand{\bm}[1]{\boldsymbol{#1}}
\newcommand{\mSigma}{\mathsf{\Sigma}}
\newcommand{\mB}{\mathsf{B}}
\newcommand{\smallocite}[1]{{\small\ocite{#1}}}
% \newcommand{\bm}[1]{\boldsymbol{#1}}
\newcommand{\dif}[1]{\text{d}{#1}}
\newcommand{\D}[1]{\text{d}{#1}}
\newcommand{\trace}[1]{\textup{trace}{#1}}

@@

The integral may often be expressed as
\begin{align}
\label{eqn:defn_mu}
\mu:= \Ex[f(\boldsymbol{X})] = \int_{[0,1]^d} f(\vx)\, \dif\vx, 
\end{align}
where $f:[0,1]^d \to \reals$ is the integrand, and $\boldsymbol{X} \sim \mathcal{U}[0,1]^d$.  The process of transforming the original integral into the form of \eqref{eqn:defn_mu} is not addressed here. See \cite[Section 2.11]{DicEtal14a} 
for a discussion of variable transformations. The cubature may be an affine function of integrand values:
\begin{align}
\label{eqn:defn_hmu}  % remove this
\hmu := w_0 + \sum_{i=1}^{n} f(\vx_i) w_i,
\end{align}
where the weights, $w_0$, and  $\vw = (w_i)_{i=1}^n \in \reals^n$, and the nodes, $\{\vx_i\}_{i=1}^n \subset [0,1]^d$, are chosen to make the error, $\abs{\mu - \hmu}$, small. The integration domain $[0,1]^d$ is convenient for the low discrepancy node sets \citep{DicEtal14a,SloJoe94} that we use.  The nodes are assumed to be deterministic.

Users of cubature algorithms typically want the error to be no greater than their specified error tolerance, denoted by $\varepsilon$.  That is, they want
\begin{align}
\label{eqn:err_crit} 
\abs{\mu - \hmu} \leq \errtol .
\end{align}
Some stopping criteria for choosing $n$ are heuristic.  Rigorous algorithms satisfying \eqref{eqn:err_crit}  typically require strong a priori assumptions about the integrand, such as an upper bound on its variance (for simple Monte Carlo) or total variation (for quasi-Monte Carlo).  We take a Bayesian approach by constructing a stopping criterion that is based on a credible interval.  We build upon the work of \cite{Dia88a}, \cite{OHa91a}, \cite{Rit00a}, \cite{RasGha03a}, \cite{BriEtal18a}, and others.  Our algorithm is an example of \emph{probabilistic numerics}.

@@

\cite{Hic17a} compares different approaches to cubature error analysis depending on whether the rule is deterministic or random and whether the integrand is assumed to be deterministic or random.  Error analysis that assumes a deterministic integrand lying in a Banach space leads to an error bound that is typically impractical for deciding how large $n$ must be to satisfy \eqref{eqn:err_crit}.  The deterministic error bound includes a (semi-) norm of the integrand, often called the variation, which is often more complex to compute than the original integral.

\cite{HicJim16a} and \cite{JimHic16a} have developed stopping criteria for cubature rules based on low discrepancy nodes by tracking the decay of the discrete Fourier coefficients of the integrand.  The algorithm proposed here also relies on discrete Fourier coefficients, but in a different way.  Although we only explore automatic Bayesian cubature for absolute error tolerances, the recent work by Hickernell, Jim\'enez-Rugama, and Li~\cite{HicEtal17a} suggests how one might accommodate more general error criteria, such as relative error tolerances.

@@

We assume that the integrand, $f$, is an instance of a Gaussian stochastic process  i.e., $f \sim \mathcal{GP}(m,s^2 C_\vtheta)$  \citep{Dia88a,OHa91a,Rit00a,RasGha03a,BriEtal18a}.  Specifically, $f$ is a real-valued random function with constant mean $m$ and covariance kernel $s^2C_\vtheta$:
\begin{gather*}
        m = \Ex[f(\vx)] \qquad \forall \vx \in \reals^d, \\
        \Ex\{[f(\vt) - m][f(\vx) -m]\} = s^2 C_\vtheta(\vt,\vx) \qquad \forall \vt, \vx \in \reals^d.
\end{gather*}
Here $s$ is a positive scale factor, and $C_\vtheta: [0,1]^d \times [0,1]^d \to \mathbb{R} $ is a symmetric, positive-definite function and parameterized by the vector $\vtheta$:
\begin{multline} \label{FJH:eq:CondPosDef}
\mC_\vtheta^T = \mC_\vtheta,  \; \va^T \mC_\vtheta \va > 0,  \text{ where }  \mC_\vtheta = \left(  C_\vtheta(\vx_i,\vx_j)  \right)_{i,j=1}^n,\\
 \forall \va \ne 0, \;
 n\in \mathbb{N}, \; \text{ distinct }\vx_1, \ldots, \vx_n \in [0,1]^d.
\end{multline}
Procedures for estimating or integrating out the hyperparameters $m$, $s$, and $\vtheta$ are explained later in this section.

@@

\begin{lemma} \citep[(A.6), (A.11--13)]{RasWil06a} \label{thrm:condDist} If $\vY = (\vY_1, \vY_2)^T \sim \calN (\vm,\mSigma)$, where $\vY_1$ and $\vY_2$ are random vectors of arbitrary length, and 
	\begin{gather*}
	\vm = \begin{pmatrix} \vm_1 \\ \vm_2 \end{pmatrix} = \begin{pmatrix} \Ex(\vY_1) \\ \Ex(\vY_2) \end{pmatrix}, \\
	\mSigma = \begin{pmatrix}
	\mSigma_{11} & \mSigma_{21}^T \\ 	\mSigma_{21} & \mSigma_{22}
	\end{pmatrix} =
	\begin{pmatrix}
	\var(\vY_{1}) & \cov(\vY_{1}, \vY_2) \\ 	\cov(\vY_2,\vY_{1}) & \var(\vY_{2}),
	\end{pmatrix} 
	\end{gather*}
	then 
	\begin{align*}
	\vY_1 \vert \vY_2 \; \sim \; \calN \bigl(\vm_1 + \mSigma_{21}^T \mSigma_{22}^{-1}(\vY_2 - \vm_2),  \; \mSigma_{11} - \mSigma_{21}^T \mSigma_{22}^{-1} \mSigma_{21} \bigr).
	\end{align*}
Moreover, the inverse of the matrix $\mSigma$ may be partitioned as
\begin{gather*}
\mSigma^{-1} = \begin{pmatrix} \mA_{11} & \mA_{21}^T \\ \mA_{21} & \mA_{22} \end{pmatrix}, \\
\mA_{11} = (\mSigma_{11} - \mSigma_{12} \mSigma_{22}^{-1} \mSigma_{21})^{-1}, \qquad 
\mA_{21} = -  \mSigma_{22}^{-1} \mSigma_{21} \mA_{11}, \\ 
\mA_{22} = \mSigma_{22}^{-1} + \mSigma_{22}^{-1} \mSigma_{21} \mA_{11} \mSigma_{21}^T \mSigma_{22}^{-1}.
\end{gather*}

@@

Let $\mA = \mC^{-1}_\vtheta$, let $\vzeta = \mA (\vy - m \vone)$, and partition $\mC_\vtheta$, $\mA$, and $\vzeta$ as
\begin{equation*}
\mC_\vtheta = \begin{pmatrix} c_{ii}  & \vC_{-i,i}^T \\  \vC_{-i,i} & \mC_{-i,-i}\end{pmatrix}, \qquad
\mA = \begin{pmatrix} a_{ii}  & \vA_{-i,i}^T \\  \vA_{-i,i} & \mA_{-i,-i}\end{pmatrix}, \qquad \vzeta = \begin{pmatrix} \zeta_i   \\  \vzeta_{-i} \end{pmatrix},
\end{equation*}
where the subscript $i$ denotes the $i^{\text{th}}$ row or column, and the subscript $-i$ denotes all rows or columns except the $i^{\text{th}}$. Following this notation, Lemma \ref{thrm:condDist} implies that 
\begin{align*}
\mathring{y}_i & = m + \vC^T_{-i,i} \mC_{-i,-i}^{-1} (\vy_{-i} -m \vone)  \\
\zeta_i  & = a_{ii}(y_i - m) + \vA_{-i,i}^T(\vy_{-i} - m \vone) \\
& = a_{ii}[(y_i - m) - \vC^T_{-i,i} \mC_{-i,-i}^{-1} (\vy_{-i} -m \vone)] \\
& = a_{ii}(y_i - \mathring{y}_i).
\end{align*}
Thus, \eqref{FJH:eq:CVA} may be re-written as 
\begin{align*} %\label{FJH:eq:CVB}
\textup{CV} = \sum_{i=1}^n \left(\frac{\zeta_i}{a_{ii}} \right)^2, \qquad \vzeta = \mC^{-1}_\vtheta(\vy - m \vone).
\end{align*}
The \emph{generalized cross-validation} criterion (GCV) replaces the $i^{\text{th}}$ diagonal element of $\mA$ in the denominator by the average diagonal element of $\mA$ \citep{CraWah79a,GolHeaWah79a,Wah90}:
\begin{equation*} 
\textup{GCV}
= \frac{\sum_{i=1}^n\zeta_i^2}{\left(\frac 1n \sum_{i=1}^n a_{ii} \right)^2} 
= \frac{(\vy - m\vone)^T \mC^{-2}_\vtheta (\vy - m \vone)}{\left(\frac 1n \trace(\mC_\vtheta^{-1}) \right)^2}.
\end{equation*}

@@

\cite{Gen93} introduced a variable transformation to transform \eqref{eqn:GaussDef} into an integral on the unit cube.  Not only does this variable transformation accommodate domains that are (semi-)infinite, it also tends to smooth out the integrand better, which expedites the cubature.  Let $\mSigma= \mL \mL^T$ be the Cholesky decomposition where $\mL = (l_{jk})_{j,k=1}^d$ is a lower triangular matrix.  Iteratively define
\allowdisplaybreaks
\begin{gather*}
\alpha_1 = \Phi(a_1), \qquad \beta_1 = \Phi(b_1),
\\
\alpha_j(x_1,\dots,x_{j-1}) = 
\Phi
\left(
\frac{1}{l_{jj}} 
\left(
a_j - \sum_{k=1}^{j-1} l_{jk} \Phi^{-1}(\alpha_k + x_k(\beta_k-\alpha_k))
\right)
\right), 
\\
\hspace{8cm} j=2,\dots,d,
\\
\beta_j(x_1,\dots,x_{j-1}) = 
\Phi
\left(
\frac{1}{l_{jj}} 
\left(
b_j - \sum_{k=1}^{j-1} l_{jk} \Phi^{-1}(\alpha_k + x_k(\beta_k-\alpha_k))
\right)
\right), 
\\
\hspace{8cm} j=2,\dots,d,
\end{gather*}
\begin{equation}
\label{fGenzdef}
f_{\text{Genz}}(\vx) = \prod_{j=1}^d [\beta_j(\vx) - \alpha_j(\vx)],
\end{equation}
where $\Phi$ is the univariate cumulative standard Gaussian distribution function.  Then, $\mu = \int_{[0,1]^{d'-1}} f_{\text{Genz}}(\vx) \, \dvx$. This approach transforms a $d'$ dimensional integral into a $d=d'-1$ dimensional integral.

@@

We use the following parameter values in the simulation: 
\begin{align*}
d' = 3, \quad \va = \begin{pmatrix}
-6 \\ -2 \\ -2
\end{pmatrix}, \quad 
\vb = \begin{pmatrix}
5 \\ 2 \\ 1
\end{pmatrix} , \quad 
\mL = \begin{pmatrix}
4 & 1 & 1 \\ 0 & 1 & 0.5 \\ 0 & 0 & 0.25
\end{pmatrix}.
\end{align*}
The node sets are randomly scrambled Sobol' points \citep{DicEtal14a,DicPil10a}.  The results are for 400 randomly chosen $\varepsilon$ in the interval $[10^{-5}, 10^{-2}]$ as shown in \figref{fig:MVN_Metern_d2b2}. In each run, the nodes are randomly scrambled.  The empirical Bayes credible intervals are used for stopping criteria.  We  observe that the algorithm meets the error criterion 95\% of the time even though we used 99\% credible intervals. One possible explanation is that the matrix inversions in the algorithm are ill-conditioned leading to numerical inaccuracies.  Another possible explanation is that this Mat\'ern covariance kernel is not a good match for the integrand.
%with Intel i7 3630QM and 16GB RAM memory
%On our test computer,

@@

An example of $64$ nodes is given in \figref{latticefig}.  The even coverage of the unit cube is ensured by a well chosen generating vector, $\vh$.  The choice of generating vector is typically done offline by computer search.  See \cite{HicNie03a}  and \cite{DicEtal14a} for more on extensible integration lattices.
\begin{figure}[htp]
	\centering
	\includegraphics[height=5cm]{ShiftedLatticePoints}
	\caption{Example of a shifted integration lattice node set  in $d=2$. 
	This figure can be reproduced using \code{PlotPoints.m} in GAIL} \label{latticefig}
\end{figure}

@@

A family of shift-invariant kernels is constructed via even degree Bernoulli polynomials:
\begin{multline}
\label{the_kernel_eqn_bernoulli}
K_\vtheta(\vx) =
\prod_{l=1}^d \biggl[
1 - (-1)^{r} \eta B_{2r}( x_l) \biggr], \\  
\forall \vt,\vx \in [0,1]^d, \  \vtheta = (r,\eta), \ r \in \naturals, \ \eta > 0.
\end{multline}
\end{subequations}
Symmetric, periodic, positive definite kernels of this form appear in  \cite{Hic96a} and  \cite{DicEtal14a}.  Bernoulli polynomials are described in \cite[Chapter 24]{OlvEtal10a}.

@@

Lattice cubature rules are known to have convergence rates that depend on the smoothness of the integrands, but that are rather independent of the choice of the integration lattice \citep{DicEtal14a}.  Thus, we expect integration lattice node sets to perform well regardless of the smoothness of the covariance kernel.  The bigger concern is whether the derivatives of the integrand are as smooth as the covariance kernel implies.  This topic is touched upon again in Section  \ref{period_var_tx}.

@@

We have implemented the fast adaptive Bayesian cubature algorithm in MATLAB as part of the Guaranteed Adaptive Integration Library (GAIL) \citep{ChoEtal19a} as  \code{cubBayesLattice\_g}. This algorithm uses the covariance kernel defined in \eqref{eqn:ourkernel} with  $r=1$ and $2$, and the periodizing variable transforms in  \secref{period_var_tx}.  The rank-1 lattice node generator is taken from \cite{Nuy17a}  (\code{exod2\_base2\_m20}).

@@

Three integrals were evaluated using the GAIL algorithm \code{cubBayesLattice\_g}:  a multivariate Gaussian  probability, Keister's example, and an option pricing example. 
These three integrands are defined below.
The sequences $\{\vx_i\}_{i=1}^\infty$ are the randomly shifted lattice node sequences supplied by GAIL. 
For each integral and each of our stopping criteria---empirical Bayes, full Bayes, and generalized cross-validation---our algorithm was run for $400$ different randomly chosen error tolerances. The error tolerances were chosen randomly in an interval depending on the difficulty of the problem. In each run, the nodes were also randomly shifted with $\mathcal{U}[0,1]$ shifts independent of each other and the error tolerances. The accuracy of the algorithm depends mildly on the shift; there is no universally optimal shift.
For each test, the execution times are plotted against $\abs{\mu - \hmu}/\varepsilon$.  We expect $\abs{\mu - \hmu}/\varepsilon$ to be no greater than one, but hopefully not too much smaller than one, which would indicate a stopping criterion that is too conservative. Figures \ref{fig:mvn-guaranteed-MLE} to \ref{fig:optprice-guaranteed-GCV} can be reproduced using the script \break \code{cubBayesLattice\_guaranteed\_plots.m} in GAIL.

@@

@@

% BibTeX users please use one of
%\bibliographystyle{spbasic}      % basic style, author-year citations
%\bibliographystyle{spmpsci}      % mathematics and physical sciences
%\bibliographystyle{spphys}       % APS-like style for physics
%\bibliography{}   % name your BibTeX data base
\bibliographystyle{abbrvnat}
\bibliography{FJHown23,FJH23}
